---
title: "J1994 scRNAseq visualization"
author: Ludmila Danilova
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE, cache = TRUE)

library(Seurat)
library(dplyr)
library(ggplot2)
library(ggprism)
library(RColorBrewer)
library(ggpubr)
library(ggrepel)
library(data.table)
library("DT")
library(kableExtra)
library(gridExtra)
library(xlsx)

load("../code_inputs/2024-04-27_TCells-BothBatches-final.rda")

cols = colorRampPalette(brewer.pal(8, 'Set2'))(12)
names(cols) = c('CD4 CTL', 'CD4 Naive', 'CD4 Proliferating', 
                'CD4 TCM', 'CD4 TEM', 'CD8 Naive', 'CD8 Proliferating', 'CD8 TCM', 
                'CD8 TEM', 'dnT', 'MAIT', 'Treg')

# create groups with TCM split into reactive and non-reactive
createCellReact = function(gr)
{
  gr$cell_react = gr$batch_annotation
  tcm = which(gr$batch_annotation == "CD4 TCM")
  gr[tcm, "cell_react"] = paste(gr[tcm, "cell_react"], gr[tcm, "antiKRAS"], sep = "_")
  gr$cell_react = gsub("CD4 TCM_TRUE","CD4 TCM_reactive", gr$cell_react)
  gr$cell_react = gsub("CD4 TCM_FALSE", "CD4 TCM_nonreactive", gr$cell_react)
  gr
}

# plot: dot plot
getDotPlot = function(dat, title = "", lab = "Cell type")
{
    ggplot(data = dat, aes(x = features.plot, y = id, color = avg.exp.scaled)) + 
    geom_point(size = 4) +
    scale_colour_gradient2(low = "blue", high = "red", name = "Expression") +
     theme_classic()+
 #     theme_cowplot()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
      ylab(lab) +
    xlab("Marker") +
  ggtitle(title)
}

```

# Intput data

Alex Girgis did analysis of scRNAseq data from J1994 trial. There are two batches of samples:

 - Batch 1: 1 PDAC sample with four time points (1 pre (C1D1GE) and 3 post (C2D1GE, C3D1GE, C6D1GE) vaccination). Time point C2D1GE corresponds to post-vaccination time point in batch 2
 
 - Batch 2: 4 patients (1 colon (004) and 3 PDAC), two time points (pre and post vaccination)
 
After discussing with Alex, we decided to use batch 2 as discovery set, mention one colon sample, but don't emphasize. These are all KRAS mutants and we are focusing on antiKRAS reactivity. The most abundant antiKRAS reactive cells are CD4 TCM. We "validate" that finding in batch 1. And we want to show something more about CD4 TCM cells and/or KRAS reactivity in terms of marker expression.0

Data is in the Seurat_Objects folder

## Batch 2 statistics

```{r batch 2 stat, echo=FALSE,  fig.width = 5.3, fig.height=3.5, message = FALSE, warning = FALSE}
# extract meta data for batch 2
meta = seu2_final@meta.data

meta %>% select(batch_annotation, antiKRAS) %>% table()
meta %>% select(batch_annotation, antiKRAS, Timepoint) %>% table()
print("CD4 TCM cells only")
meta %>% filter(batch_annotation == "CD4 TCM") %>% select(antiKRAS, Timepoint) %>% table()
meta %>% filter(batch_annotation == "CD4 TCM") %>% select(antiKRAS, Timepoint, Patient) %>% table()
print("CD8 TEM cells only")
meta %>% filter(batch_annotation == "CD8 TEM") %>% select(antiKRAS, Timepoint) %>% table()


#DimPlot(seu2_final, reduction = "umap")
```

### UMAPs

```{r umaps, echo=FALSE,  fig.width = 10, message = FALSE, warning = FALSE}
#===============
# with batch_annotation
umaps = seu2_final@reductions$umap@cell.embeddings
colnames(umaps) = c('UMAP1', 'UMAP2')
meta = cbind(meta, umaps)

gg = ggplot(meta, aes(x = UMAP1, y = UMAP2)) + 
  geom_point(size = 0.25, color = 'grey90') + 
  geom_point(data = meta[which(!is.na(meta$batch_annotation)),], aes(color = batch_annotation), size = 0.25) + 
  theme_prism() +
  guides(colour = guide_legend(override.aes = list(size=2))) + 
  theme(axis.text = element_blank(), axis.ticks = element_blank()) + 
  scale_color_manual(values = cols)

print(gg)
#=====================
# CD4 TCM cells reactive vs not
dat = meta %>% filter(batch_annotation == "CD4 TCM")
gg_reactive = ggplot(dat, aes(x = UMAP_1, y = UMAP_2)) + 
  geom_point(size = 0.25, color = 'grey90') + 
  geom_point(data = dat[which(!is.na(dat$antiKRAS)),], aes(color = antiKRAS), size = 0.25) + 
  theme_prism() +
  guides(colour = guide_legend(override.aes = list(size=2))) + 
  theme(axis.text = element_blank(), axis.ticks = element_blank()) + 
  scale_color_manual(values = c('TRUE' = 'red', 'FALSE' = "grey90"))

print(gg_reactive)
```

## Batch 1 statistics

```{r batch 1 stat, echo=FALSE,  fig.width = 10, message = FALSE, warning = FALSE}
meta1 = seu1_final@meta.data
meta1 %>% select(batch_annotation, antiKRAS) %>% table()

meta1 %>% filter(batch_annotation == "CD4 TCM") %>% select(antiKRAS, Sample) %>% table()

```

## Find markers of reactive cells withing CD4 TCM

```{r tcm reactive markers, echo=FALSE, eval=FALSE}

# create batch_annotation + reactivity phenotype
Idents(seu2_final) = paste(seu2_final$batch_annotation, seu2_final$antiKRAS, sep = "_")

tcm_react <- FindMarkers(seu2_final, ident.1 = "CD4 TCM_TRUE", ident.2 = "CD4 TCM_FALSE", verbose = FALSE, min.diff.pct = 0.1, only.pos = T)
dim(tcm_react)
head(tcm_react, n = 10)
top_react = tcm_react %>% filter(avg_log2FC > 2 & p_val_adj < .05)%>% rownames()
print(top_react)
length(top_react)

# save results for the supplement
write.csv(tcm_react[top_react,], file = "CD4_TCM_reactive_markers_batch2_table.csv")

# subset to CD4 TCM cells only 
Idents(seu2_final) = meta$batch_annotation
tcms = subset(seu2_final, idents = "CD4 TCM")

# dot plot of markers by anti KRAS activity
Idents(tcms) = tcms$antiKRAS
dp_react2 = DotPlot(tcms, features = top_react, scale = F) +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
      ylab("anti KRAS reactivity") +
    xlab("Markers") +
  scale_colour_gradient2(low = "blue", high = "red") +
  ggtitle("Batch 2. Reactivity markers in CD4 TCM cells")

dp_react_batch2 = getDotPlot(dp_react2$data, "Batch 2. CD4 TCM KRAS reactive markers", lab = "KRAS reactive")

pdf("CD4_TCM_reactive_markers_umaps_batch2.pdf")
grid.table(round(tcm_react[top_react,],3))
#print(gg_reactive)
print(dp_react2)
 s = 1:4
 for(i in 0) print(FeaturePlot(tcms, features = top_react[s+4*i]))
dev.off()

```

## Dot plot of CD4 markers

```{r cd4 markers, echo=FALSE, eval=FALSE}
# NB do not execute because knit is failing due to too big dataset

# read in CD4 markers from Amanda
cd4_markers = toupper(unlist(read.xlsx("CD4 T cell markers of interest.xlsx", 1, header = F)))

setdiff(cd4_markers,rownames(seu2_final))

# subset to CD4  cells only 
cd4Cells = grep("CD4", names(cols), value = T)
# remove CTL cells for consistency with batch 1
# because there are no those cells in batch 1
cd4Cells = setdiff(cd4Cells, "CD4 CTL")

#  batch 2
# extract meta data for batch 2
meta = seu2_final@meta.data
Idents(seu2_final) = meta$batch_annotation
cd4 = subset(seu2_final, idents = cd4Cells)


# batch 1
# extract meta data for batch 1
meta1 = seu1_final@meta.data
Idents(seu1_final) = meta1$batch_annotation
cd4_1 = subset(seu1_final, idents = cd4Cells)
Idents(cd4_1) = factor(Idents(cd4_1), levels = rev(cd4Cells))

# subset to CD8  cells only 
cd8Cells = grep("CD8", names(cols), value = T)
#  batch 2
Idents(seu2_final) = meta$batch_annotation
cd8 = subset(seu2_final, idents = cd8Cells)
# create Idents with treatment
Idents(cd8) = paste(cd8$batch_annotation, cd8$Timepoint, sep = "_")
# and save for future analysis
#saveRDS(cd8, "cd8_cells_pre_post.rds")

# find markers to get percentage of markers in each cell type
#cd4markers_seurat = FindAllMarkers(cd4)

# dot plot of CD4 markers
# couldn't make dot plot in the form I need, so switched to ggplot
dt_cd4 = DotPlot(cd4, features = cd4_markers, group.by = NULL)
dt_cd4_1 = DotPlot(cd4_1, features = cd4_markers, group.by = NULL)
dt_cd4_noscale = DotPlot(cd4, features = cd4_markers, group.by = NULL, scale = F)
dt_cd4_1_noscale = DotPlot(cd4_1, features = cd4_markers, group.by = NULL, scale = F)

# create dot plot 
dp_cd4_batch2 = getDotPlot(dt_cd4$data, "Batch 2. CD4 markers")
dp_cd4_batch1 = getDotPlot(dt_cd4_1$data, "Batch 1. CD4 markers")
dp_cd4_batch2_noscale = getDotPlot(dt_cd4_noscale$data, "Batch 2. CD4 markers")
dp_cd4_batch1_noscle = getDotPlot(dt_cd4_1_noscale$data, "Batch 1. CD4 markers")

gr = createCellReact(cd4@meta.data)
gr1 = createCellReact(cd4_1@meta.data)
# set levels, so the groups are plotted in the specific order
lv = c("CD4 TCM_reactive", "CD4 TCM_nonreactive", "CD4 TEM","CD4 Proliferating", "CD4 Naive")

Idents(cd4) = factor(gr$cell_react, levels = lv)
Idents(cd4_1) = factor(gr1$cell_react, levels = lv)

# create data to plot for both batches
dt_cd4_noscale = DotPlot(cd4, features = cd4_markers, group.by = NULL, scale = F)
dt_cd4_1_noscale = DotPlot(cd4_1, features = cd4_markers, group.by = NULL, scale = F)
# create dotplots
pdf("dotplot_cd4_react_markers.pdf", height = 4)
  getDotPlot(dt_cd4_noscale$data, "Batch 2. CD4 markers")
  getDotPlot(dt_cd4_1_noscale$data, "Batch 1. CD4 markers")
dev.off()

```


## Validation of reactive cells withing CD4 TCM in batch 1

```{r validation reactive, echo=FALSE, eval=FALSE}
# TCM markers validation
# assign clusters from batch_annotation to find markers 
Idents(seu1_final) = meta1$batch_annotation
# check assingment
head(Idents(seu1_final))
# find markers for CD4 TCM
# all cells
tcm_markers_val = FindMarkers(seu1_final, ident.1 = "CD4 TCM", min.diff.pct = 0.01)
dim(tcm_markers_val)
# check top markers from batch 2 in these results
tcm_markers_val[top_react,]

#======================
# reactive markers validation
# create batch_annotation + reactivity phenotype
Idents(seu1_final) = paste(seu1_final$batch_annotation, seu1_final$antiKRAS, sep = "_")

tcm_react_val <- FindMarkers(seu1_final, ident.1 = "CD4 TCM_TRUE", ident.2 = "CD4 TCM_FALSE", verbose = FALSE)

tcm_react_val[top_react,]
sum(tcm_react_val[top_react,] %>% filter(p_val_adj < .05) %>% dim())

  # subset to CD4 TCM cells only and make UMAP
  Idents(seu1_final) = meta1$batch_annotation
  tcms1 = subset(seu1_final, idents = "CD4 TCM")
  Idents(tcms1) = tcms1$antiKRAS
# seurat dot plot
dp_react1 = DotPlot(tcms1, features = top_react, scale = F) +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
      ylab("anti KRAS reactivity") +
    xlab("Markers") +
  scale_colour_gradient2(low = "blue", high = "red") +
  ggtitle("Batch 1. Reactivity markers in CD4 TCM cells")

# save tables with CD4 TCM markers and reactive markers and dot plot of reactive
pdf("TCM_marker_validation_batch1.pdf")
  grid.table(round(tcm_markers_val[top,],3))
  plot.new()
  grid.table(round(tcm_react_val[top_react,],3))
  dp_react1
dev.off()

dp_react_batch1 = getDotPlot(dp_react1$data, "Batch 1. CD4 TCM KRAS reactive markers", lab = "KRAS reactive")

# dot plots in one file
pdf("dotPlots.pdf", height = 4)
  dp_cd4_batch2
  dp_cd4_batch1
  dp_cd4_batch2_noscale
  dp_cd4_batch1_noscle
  dp_react2
  dp_react1
  dp_react_batch2
  dp_react_batch1
dev.off()

```


```{r echo=FALSE}
#sessionInfo()
```
